A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-01-17 doc/large_strain.tex
update discussions around MF GMG to fit new results

Next we evaluate the performance of the proposed geometric multigrid preconditioner.
{\color{red}
Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end}
we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees
{\color{red} for the GMG preconditioner.}
On the other hand, the black-box AMG preconditioner requires many more iterations for convergence.
{\color{red}
Consequently, we can conclude that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a well behaving multigrid preconditioner for the here considered finite strain elasticity problem and that the considered smoother is suitable.
}

{\color{red}
Additionally, we compare solution times of the preconditioned iterative solvers.
Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner,
as well as to identify some bottlenecks within the former method that shall guide our further research in this direction.}
Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for polynomial degrees up to $p=5$.
For higher orders in 2D the wall-clock time {\color{red}per DoF} is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
The reason for this is the suboptimal coarse level solver within the GMG preconditioner.
Evidently, the combination of the Chebyshev smoother with the CG solver at the coarsest mesh level has unsatisfactory performance for higher polynomial degrees.
{\color{red}
Consequently, the coarse level solver is the bottleneck of the here described geometric multigrid preconditioner. Its performance shall first be improved from a linear algebra perspective and only then implementational details of the complete preconditioner shall be studied.}
One possible solution to this issue would be to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
Although untested in this work, we hypothesize that this should also improve the performance of the preconditioner for lower degrees.
A valid alternative would be to use matrix-based AMG at the coarsest level.

@@

In this contribution, we proposed and numerically investigated {\color{red}three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
{\color{red}
To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to such materials.}
{\color{red}Among the three algorithms,}
the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
{\color{red}
Given that in this case we did not utilize the specific form of the material part of the stiffness tensor, we conclude}
that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.

{\color{red}
The Roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity. Further studies are required to identify means to improve the computational intensity and performance of the matrix-free implementation.}

{\color{red}
We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner
}
with standard geometric transfer operations between each level.
The GMG was applied to heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
The numerical studies indicate that the proposed preconditioner is efficient from the perspective of linear algebra.
{\color{red}
Namely, the number of iterations in the iterative solver is almost independent on the polynomial degree of the finite element basis.
The multigrid matrix-free preconditioner also}
leads to a solution approach that is faster than the matrix-based AMG for moderately high polynomial degrees (approximately up to order five).
However, a considerable amount of time is spent in the coarse level solver of the GMG preconditioner,
{\color{red}
which is the main bottleneck of the here considered iterative solver.}
Our future work will, therefore, be focused on extending the proposed matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
We are also interested in extending these methods to near incompressible three-field formulations.

{\color{red}
Based on the results presented in this work, we can conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
the geometric multigrid preconditioner is a very competitive solution strategy, as compared to more traditional matrix-based approach.}

@@

D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.

>>>>

Next we evaluate the performance of the proposed geometric multigrid preconditioner.
{\color{red}
Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end}
we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
{\color{red} noticeable} increase in the average number of CG iterations for different polynomial degrees
{\color{red} for the GMG preconditioner.
Although not reported in this section, we identified that
the coarse level solver setup will greatly affect the performance of the preconditioner.
In particular, we can improve the number of iterations by choosing a more accurate coarse level solver,
however this leads to an overall more expensive linear solver.}
The black-box AMG preconditioner
{\color{red} demonstrate the same behavior of the number of CG iterations, however it}
requires many more iterations for convergence.
{\color{red}
Consequently, we can conclude that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid for the here considered finite strain elasticity problem and that the considered smoother is suitable.
}

{\color{red}
Additionally, we compare solution times of the preconditioned iterative solvers.
Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.}
Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
{\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
On another hand, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for engendering problems
with finite strain material models.
Extension of this study from the node level to the cluster level is beyond the scope of this work,
however we hypothesize that the matrix-free and multigrid frameworks provided by \texttt{deal.II} library would provide a competitive solution for that case as well.}

@@

In this contribution, we proposed and numerically investigated {\color{red}three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
{\color{red}
To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.}
{\color{red}Among the three algorithms,}
the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
{\color{red}
Given that in this case we did not utilize the specific form of the material part of the stiffness tensor, we conclude}
that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.

{\color{red}
The Roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth order and second order symmetric tensor
as well as single contraction of two second order tensors is performed.
Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by the \texttt{deal.II}.}

{\color{red}
We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner
}
with standard geometric transfer operations between each level.
The GMG was applied to heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
The numerical studies indicate that the proposed preconditioner
{\color{red}
leads to much fewer iterations of the iterative solver as compared to the algebraic multigrid preconditioner.
The multigrid matrix-free preconditioner also}
leads to a solution approach that is faster than the matrix-based AMG {\color{red}for all polynomial degrees studied here.
Importantly the solution wall time per degree of freedom is close to being constant.
On the contrary, the matrix-based implementation with AMG becomes prohibitively expensive for higher order bases.
We conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
the geometric multigrid preconditioner is a very competitive solution strategy, as compared to more traditional matrix-based approach.
Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $FE^2$ homogenization
as well studying its behavior on the cluster level as opposed to node level, considered here.}

@@

D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.
